Making Sequential Programs Deterministic

### 2019-04-04 design.tex
Edits in 2.3.1

    \caption{Type well formedness}
\todo{What if $\beta$ has \<List> within?  Be sure everything works in that case.}

@@

\Cref{type-validity} shows the type well formedness
%\todo{The caption says
%  ``type well formedness''.  Choose one term and use it consistently
%  throughout.}
rules of \ourTypeSystem. 
(For brevity, we use ``valid'' synonymously with ``well-formed'' and
``invalid'' synonymously with ``non-well-formed''.)
As stated in the rule \rulename{noncollection}, the types \<Det> and \<NonDet>
can be written on any non-collection Java basetype.

@@

The element type of a \<NonDet> collection must be \<NonDet>.
Otherwise, the type qualifier on a collection's element must be a subtype or equal to
the collection's type qualifier (\rulename{collection} rules of
\cref{type-validity}). Note that the \rulename{ordernondet} versions also handle \<Det>,
because $\<Det> \sqsubseteq \<OrderNonDet>$.
\Cref{fig:determinism-collections} restates the \rulename{collection} rules
of
\cref{type-validity}.

@@

It is possible to store an expression of type \<Det> in a \<NonDet>
collection, but all elements retrieved from a \<NonDet> collection have
type \<NonDet>.  If the type \ptype{NonDet List}{Det String} were
permitted, the following type hole would exist:

\begin{Verbatim}[commandchars=\\\{\}]
\lptype{Det    List}{Det String} ddlist
\lptype{NonDet List}{Det String} ndlist = ddlist   // permitted by subtyping rules
NonDet Int ni = ...
ndlist[ni] = "anywhere"
\end{Verbatim}

\noindent
The deterministic string \<"anywhere"> is placed in the list at an
arbitrary (that is, nondeterministic) index.  This must be prohibited,
because it unsoundly permits the deterministic list \<ddlist> to differ
from execution to execution.

@@

    $\infer[\rulename{ordernondet polydown}]
    {\down{\tau} = \<Det> \ \beta}
    {\tau = \<OrderNonDet> \ \beta}
        & \wellformed{\tau}
    }$

>>>>

    \caption{Type well-formedness}
\todo{What if $\beta$ has \<List> within?  Be sure everything works in that case.}

@@

% "invalid" is easier to read than "non-well-formed".
\Cref{type-validity} shows the type well-formedness, or type validity,
rules of \ourTypeSystem.

@@

The element type of a \<NonDet> collection must be \<NonDet>.
Otherwise, the type qualifier on a collection's element must be a subtype or equal to
the collection's type qualifier (\rulename{collection} rules of
\cref{type-validity}). Note that the \rulename{ordernondet} rules also
permit \<Det> collections,
because $\<Det> \sqsubseteq \<OrderNonDet>$.
\Cref{fig:determinism-collections} restates the \rulename{collection} rules
of
\cref{type-validity}.

@@

It is possible to store an expression of type \<Det> in a \<NonDet>
collection, but any value retrieved from a \<NonDet> collection has
type \<NonDet>.

If the type \ptype{NonDet List}{Det String} were
permitted, the following type hole would exist:

\begin{Verbatim}[commandchars=\\\{\}]
\lptype{Det    List}{Det String} ddlist
\lptype{NonDet List}{Det String} ndlist = ddlist   // assignment permitted by subtyping rules
NonDet Int ni = ...
ndlist[ni] = "anywhere"
\end{Verbatim}

\noindent
In this code snippet,
the deterministic string \<"anywhere"> is placed in the list at an
arbitrary (that is, nondeterministic) index.  This must be prohibited,
because it unsoundly permits the deterministic list \<ddlist> to differ
from execution to execution.  \OurTypeSystem does so by forbidding the type
\ptype{NonDet List}{Det String}.

@@

    $\infer[\rulename{ordernondet polydown}]
    {\down{\tau} = \<Det> \ \beta}
    {\tau = \<OrderNonDet> \ \beta
        & \wellformed{\tau}
    }$
